A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-12 tex/sec6-formalisation.tex
[ tex ] Done with formalisation (for now ...)

% TODO: Define more of these ...
\DeclareUnicodeCharacter{737}{\textsuperscript{l}}
\DeclareUnicodeCharacter{8718}{\ensuremath{\blacksquare}}
\DeclareUnicodeCharacter{8759}{::}
\DeclareUnicodeCharacter{9669}{\ensuremath{\triangleleft}}
\DeclareUnicodeCharacter{8799}{\ensuremath{\stackrel{\scriptscriptstyle ?}{=}}}
\DeclareUnicodeCharacter{10214}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{10215}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{27E6}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{27E7}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}

@@

\section{Formalisation} \label{sec:formalisation}
\begin{enumerate}
  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
  \item type synthesis and checking -- 0.5 p
  \item Examples including STLC (PCF), application in spine form --- 1p

\end{enumerate}

\begin{figure}
  \small
  \begin{minipage}[t]{.3\linewidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{mode}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.32\linewidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[14]%
\>[14]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[14]%
\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgsD}\;\Varid{vars}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[14]%
\>[14]{}\mathbin{:}\;\Conid{TExp}\;{}\<[22]%
\>[22]{}\Varid{vars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{mode}\;{}\<[14]%
\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\linewidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[17]%
\>[17]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{rules}\;{}\<[17]%
\>[17]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{ConD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}

  \caption{Definition of bidirectional binding signature}  
\end{figure}

{\small\begin{hscode}\SaveRestoreHook

@@

\column{7}{@{}>{\hspre}l<{\hspost}@{}}%

@@

@@

\column{24}{@{}>{\hspre}l<{\hspost}@{}}%

>>>>

% TODO: Define more of these ...
\DeclareUnicodeCharacter{737}{\textsuperscript{l}}
\DeclareUnicodeCharacter{8718}{\ensuremath{\blacksquare}}
\DeclareUnicodeCharacter{8759}{::}
\DeclareUnicodeCharacter{9669}{\ensuremath{\triangleleft}}
\DeclareUnicodeCharacter{8799}{\ensuremath{\stackrel{\scriptscriptstyle ?}{=}}}
\DeclareUnicodeCharacter{10214}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{10215}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{27E6}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{27E7}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}
\DeclareUnicodeCharacter{2218}{\ensuremath{\circ}}

@@

\section{Formalisation} \label{sec:formalisation}
As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda.
While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could pose some difficulty.
If the reader is comfortable with the informal presentation and assured by the existence of its formalisation, they may feel free to skip this section.

\paragraph{Revisiting language formalisation frameworks}
Unlike prior frameworks~\citep{Allais2021,Fiore2022,Ahrens2022} that have primarily focused on meta-properties centred around substitution for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
The formal definitions of extrinsic typing are more complex than their intrinsic counterparts.
For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
For our generic definitions, take the formal type-theoretic definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
Its intrinsic definition consists of of just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.

\paragraph{Category-theoretic analysis of well-typed terms}
The difference between intrinsic typing and extrinsic typing could be partly manifested by its categorical analysis.
\citet{Fiore1999}{'s} theory of abstract syntax and variable binding forms the foundation of \citet{Fiore2022}'s framework and inspires frameworks by \citet{Allais2021,Ahrens2022}.
Let us take a closer look of their idea.
The set of (untyped) abstract syntax trees for a language can be understood as
\begin{enumerate*}
  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs. 
  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
\end{enumerate*}
Substitution is modelled as a monoid multiplication with the notion of strength and a suitable monoidal structure on the category of presheaves.
To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.

Such an initial algebra can be constructed as the least fixpoint of a sequence of objects defined via iteratively applying the signature functor by the Initial Algebra Theorem~\cite{Trnkova1975}.

\paragraph{Type-theoretic construction of well-typed terms}
Fortunately, in type theory, constructing the initial algebra of well-typed terms boils down to defining an inductive type with a few constructors that align primitive rules (such as $\Rule{Var}$) and a rule schema; term traversal can be defined by pattern matching as usual for the inductive type~\citep{Fiore2022}.

\paragraph{Lack of a theory for extrinsic typing}
While the theory of \citet{Fiore1999} and others provide significant insights, they do not consider extrinsic typing.
We find inspiration in the interpretation of structural induction as algebras for an endofunctor on the category of predicates over a base category, as put forward by \citet{Hermida1998}.
From this perspective, we interpret extrinsic typing as a signature `functor' on the category of predicates consisting of a $\N$-indexed family $X$ of sets and another family $Y$ of sets over $X$; constructing typing derivations as its initial algebra over well-scoped terms.
Our formal construction appears novel and we believe that a theoretic understanding of our construction and its connection to intrinsic typing as a bridge between raw terms and well-typed terms deserve some attention.
However, carrying out a category-theoretic analysis requires expertise in category theory, which goes beyond the scope of this paper.

Therefore, in this section, we will primarily provide an overview of our design and offer intuitive explanations.
We will discuss the construction of simple types in \Cref{subsec:formal-simple-types}, raw terms in~\Cref{subsec:formal-raw-terms}, and bidirectional typing rules in \Cref{subsec:formal-bidirectional-typing}.
Since the formal proofs largely reflect their informal presentation, we will limit our illustration to the 'if' part of \Cref{lem:soundness-completeness} as an example in \Cref{subsec:formal-proofs}.

\subsection{Defining Simple Types}\label{subsec:formal-simple-types}

\begin{figure}
  \codefigure
  \small
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{decEq}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{DecEq}\;\Conid{Op}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Op}\;\Varid{→}\;\Conid{ℕ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;\Conid{SigD}\;\Varid{→}\;\Conid{Set}\;\Varid{→}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Conid{X}\;\hat{}\;(\Conid{D}\;\Varid{.ar}\;\Varid{i}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{Ty}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Ξ}\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{op}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Ty}\;\Conid{Ξ})\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Simple types in \Agda}
  \label{fig:formal-simple-type}
\end{figure}
We start with the formal definition of signatures and simple types (\Cref{fig:formal-simple-type}), which parallels its informal counterpart (\Cref{def:simple-signature}), with the exception of \ensuremath{\Varid{⟦\char95 ⟧}}.
To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of \ensuremath{\Conid{Ty}}.
Thus, types specified by a signature can be defined using two constructors--- \ensuremath{\text{\textasciigrave}\anonymous }\ and \ensuremath{\Varid{op}}---where a variable is represented by \ensuremath{\text{\textasciigrave}\;\Varid{n}} for some \ensuremath{\Varid{n}} less than \ensuremath{\Conid{Ξ}}, and each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} is a pair of an operation symbol \ensuremath{\Varid{i}} and a list \ensuremath{\Varid{ts}} of length $n$.

From a categorical standpoint, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} acts as a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
The type \ensuremath{\Conid{Ty}} is the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.

Finally, to account for indices when defining raw terms and bidirectional typing derivations by a signature, we define a corresponding signature functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} for a (bidirectional) binding signature.

\subsection{Defining Raw Terms}\label{subsec:formal-raw-terms}
As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook

@@

\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\>[3]{}\Keyword{where}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}
To construct well-scoped raw terms indexed by a list $V$ of free variables, we introduce another signature functor mapping \ensuremath{\Conid{Fam}} to \ensuremath{\Conid{Fam}}, where \ensuremath{\Conid{Fam}} is the family of sets defined as \ensuremath{\Conid{ℕ}\;\Varid{→}\;\Conid{Set}}, indexed by a natural number indicating the number of free variables.
This signature endofunctor for raw term construction resembles the one for defining simple types.
However, we define four distinct functors instead of defining one directly.
These are \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} for an extension context \ensuremath{\Conid{Δ}}, \ensuremath{\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}} for an argument list \ensuremath{\Conid{Ds}}, \ensuremath{\Varid{⟦}\;\Varid{ι}\;\Conid{Ds}\;\anonymous \;\Varid{⟧ᶜ}} for an operation (\ensuremath{\Varid{ι}\;\Conid{Ds}\;\Conid{A}}), and \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}} for a binding signature $D$.
Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by $\erase{\Delta}$ variables by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.
\begin{figure}
  \codefigure
\begin{minipage}[t]{.33\textwidth}
\begin{hscode}\SaveRestoreHook
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ}\;{}\<[19]%
\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
\>[B]{}\Varid{⟦}\;\Conid{Δ}\;{}\<[17]%
\>[17]{}\Varid{⟧ᵃ}\;{}\<[22]%
\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
\>[27]{}\mathrel{=}\;{}\<[E]%
\>[3]{}\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n}){}\<[E]%
\>[8]{}\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ}\;{}\<[19]%
\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
\>[22]{}\anonymous \;\anonymous \;{}\<[27]%
\>[27]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
\>[B]{}\Varid{⟦}\;(\Conid{Δ}\;\Varid{⊢}\;\Conid{A})\;\Varid{∷}\;\Conid{Ds}\;{}\<[17]%
\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
\>[27]{}\mathrel{=}\;{}\<[E]%
\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Varid{n}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Varid{n}{}\<[E]%
\end{minipage}
\begin{minipage}[t]{.3\textwidth}
\begin{hscode}\SaveRestoreHook

@@

\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%

@@

\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{OpD}\;{}\<[15]%
\>[15]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
\>[B]{}\Varid{⟦}\;\Varid{ι}\;\Conid{Ds}\;\anonymous \;{}\<[11]%
\>[11]{}\Varid{⟧ᶜ}\;{}\<[15]%
\>[15]{}\Conid{X}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}{}\<[E]%
\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{SigD}\;{}\<[16]%
\>[16]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[10]%
\>[10]{}\Varid{⟧}\;{}\<[15]%
\>[15]{}\Conid{X}\;\Varid{n}\;{}\<[20]%
\>[20]{}\mathrel{=}\;{}\<[E]%
\>[3]{}\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Varid{n}{}\<[E]%
\end{minipage}
\begin{minipage}[t]{.33\textwidth}
\begin{hscode}\SaveRestoreHook
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{data}\;\Conid{Raw}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Varid{n}\;{}\<[33]%
\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{Ty}\;{}\<[23]%
\>[23]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
\>[3]{}\Varid{op}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
\end{minipage}
\caption{Signature functor for raw terms and raw terms in \Agda}
\label{fig:formal-raw-term-functor}
\label{fig:formal-raw-terms}
\end{figure}

The inductive type of raw terms, indexed by a list of free variables, is defined as shown in \Cref{fig:formal-raw-terms}.
This mirrors our informal definition (\Cref{fig:raw-terms}), where \ensuremath{\Conid{A}\;\Varid{∋}\;\Varid{t}} corresponds to the raw term $t \annotate A$.
These definitions align closely with those found in other referenced frameworks.

\subsection{Defining Extrinsic Bidirectional Typing Derivations}\label{subsec:formal-bidirectional-typing}
Formally defining extrinsic derivations is arguably the most complex aspect of our formalisation.
This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term.
This definition involves another signature endofunctor, requiring consistency with the functor used for constructing raw terms.

We begin with the family of sets on which the functor acts.
A typing judgement $\Gamma \vdash_{\Sigma, \Omega} t :^{\dir{d}} A$ can be viewed as a predicate over raw terms with four indices: a context, a $\N$-indexed family of sets, a mode, and a type.
The type of this predicate can be generalised to a family of sets over the $\N$-indexed family of sets, as follows:
{\small
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Fam}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{→}\;\Conid{Set}\;\anonymous {}\<[E]%
\\
\>[B]{}\Conid{Fam}\;\Conid{X}\;{}\<[8]%
\>[8]{}\mathrel{=}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Mode}\;\Varid{→}\;\Conid{Ty}\;\Varid{→}\;\Conid{Set}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
Observe that the second index \ensuremath{\Conid{X}} depends on the length of the first index \ensuremath{\Conid{Γ}}.
This is because a typing derivation for a raw term $V \vdash_{\Sigma, \erase{\Omega}} t$ requires the same number of variables as in $V$.

In a similar vein, we define four functors for the construction of bidirectional typing derivations, albeit in a top-down manner.
For an endofunctor on predicates, each functor must act as a \emph{lifting} of the functor on its domain, illustrated below.
The signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature is defined as follows:
{\small
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[6]%
\>[6]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{SigD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{o}\;\Varid{,}\;\Varid{ts})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{ts}\;\Varid{d}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
This functor maps a predicate \ensuremath{\Conid{Y}} over \ensuremath{\Conid{X}} into a predicate \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}} over \ensuremath{\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}}.
Here, \ensuremath{\Varid{erase}\;\Conid{D}} represents the mode erasure of \ensuremath{\Conid{D}}, and the prefix \ensuremath{\Conid{R}} in \ensuremath{\Conid{R.⟦\char95 ⟧}} indicates the signature endofunctor used in raw term construction.
